A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-22 text/formal-defnition.tex
Revise.
Minor edits.

\section{Formal Definition of \LMD \label{sec:formal-definition}}

@@

{\small
\begin{align*}
    % \textrm{Type variables}  &   &                          & X,Y,Z                                                                                                      \\
    % \textrm{Variables}       &   &                          & x,y,z                                                                                                      \\
    % \textrm{Stage variables} &   &                          & \alpha,\beta,\gamma                                                                                        \\
    % \textrm{Stage}           &   &                          & A,B,C                                                                                                      \\
    \textrm{kinds}             &  & K,J,I,H,G                & ::= * \mid \Pi x:\tau.K                                                           \\
    \textrm{types}             &  & \tau,\sigma,\rho,\pi,\xi & ::= X \mid \Pi x:\tau.\tau \mid \tau\ M \mid \TW_{\alpha} \tau \mid \F\alpha.\tau \\
    %     \textrm{Constants}       &   &                          & c                                                                                                          \\
    \textrm{terms}             &  & M,N,L,O,P                & ::= c \mid x \mid \lambda x:\tau.M\ \mid M\ M \mid \TB_\alpha M                   \\
                               &  &                          & \ \ \ \ \mid \TBL_\alpha M \mid \Lambda\alpha.M \mid M\ A \mid \%_\alpha M        \\
    \textrm{signatures}         &  & \Sigma                   & ::= \emptyset \mid \Sigma, X::K \mid \Sigma, c:\tau                               \\
    \textrm{type environments} &  & \Gamma                   & ::= \emptyset \mid  \Gamma,x:\tau @A                                              \\
\end{align*}
}
% \AI{Add $M\,\alpha$ to terms.}

@@

We adopt the tradition of \LLF-like systems, where types of constants and kinds of type-level constants are globally declared in a signature $\Sigma$, which
is a sequence of declarations of the form $c:\tau$ and $X:K$.
For example, when we use Boolean in \LMD, $\Sigma$ includes $\B :: *, \textrm{true}:\B, \textrm{false}:\B$.
Type environments are sequences of triples of a variable, its type, and its stage.
We write \(\textit{dom}(\Sigma)\) and \(\textit{dom}(\Gamma)\)
for the set of (type-level) constants declared in \(\Sigma\) and \(\Gamma\), respectively.   As in other multi-stage calculi~\cite{Taha03,Tsukada,Hanada2014},
a variable declaration is associated with a stage
so that a variable can be referenced only at the declared stage.
Constants and type-level constants are not associated with stages;
so, we can use them at any stage.
We define well-formed signature and well-formed type environments later.

@@

Next, we define full reduction for \LMD.
Before giving the definition of reduction, we define six kinds of substitutions.
Substitution $M[x\mapsto N], \tau[x \mapsto N]$ and $K[x \mapsto N]$ are
ordinary capture-avoiding substitution of
term $N$ for $x$ in term $M$, type $\tau$, and kind $K$, respectively,
and we omit their definitions here.
Substitution $M[\alpha \mapsto A], \tau [\alpha \mapsto A], K[\alpha \mapsto A]$ and $B[\alpha\mapsto A]$ are
substitutions of stage $A$ for stage variable $\alpha$ in
term $N$ for $x$ in term $M$, type $\tau$, kind $K$, and stage $B$, respectively.
We also define substitutions of a stage or a term for variables in type environment $\G$.
We show representative cases below.
%
{\small
\begin{align*}
    (\lambda x:\tau.M)[\alpha \mapsto A] & = \lambda x:(\tau[\alpha \mapsto A]).(M[\alpha \mapsto A])                                  \\
    (M\ B)[\alpha \mapsto A]             & = (M[\alpha \mapsto A])\ B[\alpha\mapsto A]                                                 \\
    (\TB_\beta M)[\alpha \mapsto A]      & = \TB_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                          \\
    (\TBL_\beta M)[\alpha \mapsto A]     & = \TBL_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                         \\
    (\%_\beta M)[\alpha \mapsto A]       & = \%_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                           \\
    (\beta B)[\alpha \mapsto A]          & = \beta (B[\alpha\mapsto A])                               & (\text{if } \alpha \neq \beta) \\
    (\beta B)[\alpha \mapsto A]          & = A (B[\alpha\mapsto A])                                   & (\text{if } \alpha = \beta)
\end{align*}
}
Here, $\TB_{\alpha_1\cdots\alpha_n} M$,
$\TBL_{\alpha_1\cdots\alpha_n} M$, and $\%_{\alpha_1\cdots\alpha_n} M$
$(n \geq 0)$ stand for $\TB_{\alpha_1} \cdots \TB_{\alpha_n} M$,
$\TBL_{\alpha_1}\cdots \TBL_{\alpha_n} M$, and
$\%_{\alpha_1}\cdots \%_{\alpha_n} M$, respectively.  In particular,
$\TB_{\varepsilon} M = \TBL_{\varepsilon} M = \%_{\varepsilon} M = M$.

\begin{definition}[Reduction]
    The relations $M \longrightarrow_\beta N$, $M \longrightarrow_\blacklozenge N$, and $M \longrightarrow_\Lambda N$
    are the least compatible relations closed under the rules below.
%    Congruence rules are omitted from the definition.
{\small
    \begin{align*}
         (\lambda x:\tau.M) N & \longrightarrow_\beta M[x \mapsto N]         \\
         \TBL_\alpha \TB_\alpha M & \longrightarrow_\blacklozenge M          \\
         (\Lambda \alpha.M)\ A & \longrightarrow_\Lambda M[\alpha \mapsto A]
    \end{align*}
  }    
\end{definition}
We write $ M \longrightarrow M'$ iff $ M \longrightarrow_\beta M'$,
$ M \longrightarrow_\blacklozenge M'$, or
$ M \longrightarrow_\Lambda M'$ and we call $\longrightarrow_\beta$,
$\longrightarrow_\blacklozenge$, and $\longrightarrow_\Lambda$
$\beta$-reduction, $\blacklozenge$-reduction, and $\Lambda$-reduction,
respectively.
$M \longrightarrow^* N$ means that there is a sequence of reduction $\longrightarrow$ whose length is greater than or equal to 0.

@@

\begin{figure}
  \begin{center}
    \infrule[\TConst]{c:\tau \in \Sigma}{\G \V c:\tau @A} \hfil
    \infrule[\TVar]{x:\tau @A \in \G}{\G \V x:\tau @A} \\[2mm]
    \infrule[\TAbs]{\G\V \sigma::*@A\andalso\G,x:\sigma@A\V M:\tau @A}{\G\V(\lambda (x:\sigma).M):(\Pi (x:\sigma).\tau)@A} \\[2mm]
    \infrule[\TApp]{\G\V M:(\Pi (x:\sigma).\tau)@A \andalso \G\V N:\sigma@A}{\G\V M\ N : \tau[x\mapsto N]@A} \\[2mm]
    \infrule[\TConv]{\G\V M:\tau @A \andalso \G\V \tau\equiv \sigma :: K@A}{\G\V M:\sigma@A} \\[2mm]
    \infrule[\TTB]{\G\V M:\tau @{A\alpha}}{\G\V\TB_{\alpha}M:\TW_{\alpha}\tau @A} \andalso
    \infrule[\TTBL]{\G\V M:\TW_{\alpha}\tau @A}{\G\V\TBL_{\alpha}M:\tau @{A\alpha}} \\[2mm]
    \infrule[\TGen]{\G\V M:\tau @A \andalso \alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)}{\G\V\Lambda\alpha.M:\forall\alpha.\tau @A} \\[2mm]
    \infrule[\TIns]{\G\V M:\forall\alpha.\tau @A}{\G\V M\ B:\tau[\alpha \mapsto B]@A} \andalso
    \infrule[\TCsp]{\G\V M:\tau @A}{\G\V \%_\alpha M:\tau @{A\alpha}}
    \caption{Typing Rules.}
    \label{fig:typing-rules}
  \end{center}
\end{figure}

@@

The helper function vadd$_1$ takes a stage, the length $n$ of vectors, and two quoted vectors and returns code that computes the addition of the given two vectors:
%
\newcommand{\Vpn}{\text{Vector}\ (\%_\alpha n)}
{\small
\begin{align*}
	  & \textbf{let}\ \text{vadd}_1 : \F\alpha.\Pi n:\I.\TW_\alpha\Vpn\to\TW_\alpha\Vpn\to\TW_\alpha\Vpn                                \\
	  & \hspace{7mm} = \textbf{fix}\ f.\Lambda\alpha.\lambda n:\I.\ \lambda v_1:\TW_\alpha\Vpn.\ \lambda v_2:\TW_\alpha\Vpn.            \\
	  & \hspace{14mm} \textbf{if}\ n = 0 \ \textbf{then}\ \TB_\alpha \text{nil}                                                                               \\
	  & \hspace{14mm} \textbf{else}\ \TB_\alpha (                                                                                       \\
	  & \hspace{21mm} \textbf{let}\ t_1 = \text{tail}\ (\TBL_\alpha v_1)\ \textbf{in}                                                   \\
	  & \hspace{21mm} \textbf{let}\ t_2 = \text{tail}\ (\TBL_\alpha v_2)\ \textbf{in}                                                   \\
	  & \hspace{21mm} \text{cons}\ (\text{head}\ (\TBL_\alpha v_1) + \text{head}\ (\TBL_\alpha v_2))\ (\TBL_\alpha f\ (n-1)\ t_1\ t_2))
\end{align*}
}
(Here, we assume that the type system can know whether $n=0$ when
\textbf{then}- and \textbf{else}-branches are typechecked so that
both branches can be given type \(\TW_\alpha \text{Vector }n\).)

>>>>

\section{Formal Definition of \LMD \label{sec:formal-definition}}
\label{sec:formal}

@@

{%\small
\begin{align*}
    % \textrm{Type variables}  &   &                          & X,Y,Z                                                                                                      \\
    % \textrm{Variables}       &   &                          & x,y,z                                                                                                      \\
    % \textrm{Stage variables} &   &                          & \alpha,\beta,\gamma                                                                                        \\
    % \textrm{Stage}           &   &                          & A,B,C                                                                                                      \\
    \textrm{kinds}             &  & K,J,I,H,G                & ::= * \mid \Pi x:\tau.K                                                           \\
    \textrm{types}             &  & \tau,\sigma,\rho,\pi,\xi & ::= X \mid \Pi x:\tau.\tau \mid \tau\ M \mid \TW_{\alpha} \tau \mid \F\alpha.\tau \\
    %     \textrm{Constants}       &   &                          & c                                                                                                          \\
    \textrm{terms}             &  & M,N,L,O,P                & ::= c \mid x \mid \lambda x:\tau.M\ \mid M\ M \mid \TB_\alpha M                   \\
                               &  &                          & \ \ \ \ \mid \TBL_\alpha M \mid \Lambda\alpha.M \mid M\ A \mid \%_\alpha M        \\
    \textrm{signatures}         &  & \Sigma                   & ::= \emptyset \mid \Sigma, X::K \mid \Sigma, c:\tau                               \\
    \textrm{type env.} &  & \Gamma                   & ::= \emptyset \mid  \Gamma,x:\tau @A                                              \\
\end{align*}
}
% \AI{Add $M\,\alpha$ to terms.}

@@

We adopt the tradition of \LLF-like systems, where types of constants and kinds of type-level constants are globally declared in a signature $\Sigma$, which
is a sequence of declarations of the form $c:\tau$ and $X:K$.
For example, when we use Boolean in \LMD, $\Sigma$ includes $\B :: *, \textrm{true}:\B, \textrm{false}:\B$.
Type environments are sequences of triples of a variable, its type, and its stage.
We write \(\textit{dom}(\Sigma)\) and \(\textit{dom}(\Gamma)\)
for the set of (type-level) constants declared in \(\Sigma\) and \(\Gamma\), respectively.   As in other multi-stage calculi~\cite{taha2003environment,Tsukada,Hanada2014},
a variable declaration is associated with a stage
so that a variable can be referenced only at the declared stage.
Constants and type-level constants are not associated with stages;
so, we can use them at any stage.
We define well-formed signature and well-formed type environments later.

@@

Next, we define full reduction for \LMD.
Before giving the definition of reduction, we define six kinds of substitutions.
Substitution $M[x\mapsto N], \tau[x \mapsto N]$ and $K[x \mapsto N]$ are
ordinary capture-avoiding substitution of
term $N$ for $x$ in term $M$, type $\tau$, and kind $K$, respectively,
and we omit their definitions here.
Substitution $M[\alpha \mapsto A], \tau [\alpha \mapsto A], K[\alpha \mapsto A]$ and $B[\alpha\mapsto A]$ are
substitutions of stage $A$ for stage variable $\alpha$ in
term $N$ for $x$ in term $M$, type $\tau$, kind $K$, and stage $B$, respectively.
We also define substitutions of a stage or a term for variables in type environment $\G$.
We show representative cases below.
%
{%\small
\begin{align*}
    (\lambda x:\tau.M)[\alpha \mapsto A] & = \lambda x:(\tau[\alpha \mapsto A]).(M[\alpha \mapsto A])                                  \\
    (M\ B)[\alpha \mapsto A]             & = (M[\alpha \mapsto A])\ B[\alpha\mapsto A]                                                 \\
    (\TB_\beta M)[\alpha \mapsto A]      & = \TB_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                          \\
    (\TBL_\beta M)[\alpha \mapsto A]     & = \TBL_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                         \\
    (\%_\beta M)[\alpha \mapsto A]       & = \%_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                           \\
    (\beta B)[\alpha \mapsto A]          & = \beta (B[\alpha\mapsto A])                               & (\text{if } \alpha \neq \beta) \\
    (\beta B)[\alpha \mapsto A]          & = A (B[\alpha\mapsto A])                                   & (\text{if } \alpha = \beta)
\end{align*}
}
Here, $\TB_{\alpha_1\cdots\alpha_n} M$,
$\TBL_{\alpha_1\cdots\alpha_n} M$, and $\%_{\alpha_1\cdots\alpha_n} M$
$(n \geq 0)$ stand for $\TB_{\alpha_1} \cdots \TB_{\alpha_n} M$,
$\TBL_{\alpha_1}\cdots \TBL_{\alpha_n} M$, and
$\%_{\alpha_1}\cdots \%_{\alpha_n} M$, respectively.  In particular,
$\TB_{\varepsilon} M = \TBL_{\varepsilon} M = \%_{\varepsilon} M = M$.

\begin{definition}[Reduction]
    The relations $M \longrightarrow_\beta N$, $M \longrightarrow_\blacklozenge N$, and $M \longrightarrow_\Lambda N$
    are the least compatible relations closed under the rules below.
%    Congruence rules are omitted from the definition.
{%\small
    \begin{align*}
         (\lambda x:\tau.M) N & \longrightarrow_\beta M[x \mapsto N]         \\
         \TBL_\alpha \TB_\alpha M & \longrightarrow_\blacklozenge M          \\
         (\Lambda \alpha.M)\ A & \longrightarrow_\Lambda M[\alpha \mapsto A]
    \end{align*}
  }    
\end{definition}
We write $ M \longrightarrow M'$ iff $ M \longrightarrow_\beta M'$,
$ M \longrightarrow_\blacklozenge M'$, or
$ M \longrightarrow_\Lambda M'$ and we call $\longrightarrow_\beta$,
$\longrightarrow_\blacklozenge$, and $\longrightarrow_\Lambda$
$\beta$-reduction, $\blacklozenge$-reduction, and $\Lambda$-reduction,
respectively.
$M \longrightarrow^* N$ means that there is a sequence of reduction $\longrightarrow$ whose length is greater than or equal to 0.

@@

\begin{figure}[tbp]
  \begin{center}
    \infrule[\TConst]{c:\tau \in \Sigma}{\G \V c:\tau @A} \hfil
    \infrule[\TVar]{x:\tau @A \in \G}{\G \V x:\tau @A} \\[2mm]
    \infrule[\TAbs]{\G\V \sigma::*@A\andalso\G,x:\sigma@A\V M:\tau @A}{\G\V(\lambda (x:\sigma).M):(\Pi (x:\sigma).\tau)@A} \\[2mm]
    \infrule[\TApp]{\G\V M:(\Pi (x:\sigma).\tau)@A \andalso \G\V N:\sigma@A}{\G\V M\ N : \tau[x\mapsto N]@A} \\[2mm]
    \infrule[\TConv]{\G\V M:\tau @A \andalso \G\V \tau\equiv \sigma :: K@A}{\G\V M:\sigma@A} \\[2mm]
    \infrule[\TTB]{\G\V M:\tau @{A\alpha}}{\G\V\TB_{\alpha}M:\TW_{\alpha}\tau @A} \andalso
    \infrule[\TTBL]{\G\V M:\TW_{\alpha}\tau @A}{\G\V\TBL_{\alpha}M:\tau @{A\alpha}} \\[2mm]
    \infrule[\TGen]{\G\V M:\tau @A \andalso \alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)}{\G\V\Lambda\alpha.M:\forall\alpha.\tau @A} \\[2mm]
    \infrule[\TIns]{\G\V M:\forall\alpha.\tau @A}{\G\V M\ B:\tau[\alpha \mapsto B]@A} \andalso
    \infrule[\TCsp]{\G\V M:\tau @A}{\G\V \%_\alpha M:\tau @{A\alpha}}
    \caption{Typing Rules.}
    \label{fig:typing-rules}
  \end{center}
\end{figure}

@@

The helper function vadd$_1$ takes a stage, the length $n$ of vectors, and two quoted vectors and returns code that computes the addition of the given two vectors:
%
\newcommand{\Vpn}{\text{Vector}\ (\%_\alpha n)}
\begin{align*}
	  & \textbf{let}\ \text{vadd}_1 : \F\alpha.\Pi n:\I.\TW_\alpha\Vpn\to\TW_\alpha\Vpn\to\TW_\alpha\Vpn                                \\
	  & \hspace{7mm} = \textbf{fix}\ f.\Lambda\alpha.\lambda n:\I.\ \lambda v_1:\TW_\alpha\Vpn.\ \lambda v_2:\TW_\alpha\Vpn.            \\
	  & \hspace{14mm} \textbf{if}\ n = 0 \ \textbf{then}\ \TB_\alpha \text{nil}                                                                               \\
	  & \hspace{14mm} \textbf{else}\ \TB_\alpha (                                                                                       \\
	  & \hspace{21mm} \textbf{let}\ t_1 = \text{tail}\ (\TBL_\alpha v_1)\ \textbf{in}                                                   \\
	  & \hspace{21mm} \textbf{let}\ t_2 = \text{tail}\ (\TBL_\alpha v_2)\ \textbf{in}                                                   \\
	  & \hspace{21mm} \text{cons}\ (\text{head}\ (\TBL_\alpha v_1) + \text{head}\ (\TBL_\alpha v_2))\ (\TBL_\alpha f\ (n-1)\ t_1\ t_2))
\end{align*}
(Here, we assume that the type system can know whether $n=0$ when
\textbf{then}- and \textbf{else}-branches are typechecked so that
both branches can be given type \(\TW_\alpha \text{Vector }n\).)
